Analysis of competitive surfing tournaments with generalized Bradley-Terry

### 2021-08-22 inst/surfing_jsa.tex
tweaks

Figure~\ref{piechartmax} shows the maximum likelihood estimate for the
23 competitors' strengths, which appears to show a wide range from
Florence at about 17.5\% down to Coffin at about 1.5\%.  The {\tt
  hyper2} software may be used to formally assess one plausible null:
that the competitors all have equal skill, that is $H_0\colon
p_\mathrm{Andino} = p_\mathrm{Bourez}=\ldots=
p_\mathrm{Wright}=\frac{1}{23}$, and any differences in placings are
due to random variation.  The Method of Support \cite{edwards1992}
rejects the null (in favour of the evaluate) at $p=4\times 10^{-5}$.
There is thus strong evidence to suggest that the competitors are
indeed of different skills and the result actually reflects genuine
differences in abilities, rather than the result of random chance.
Observe that the competitors' points (that is, the final column of
table~\ref{resultstable}) cannot be used for this type of
probabalistic analysis: the points awarded are arbitrary and changing
the points system would change the points scored but not affect the
competitors' behaviour.

@@

The statistical method described here can test many different
suggestions drawn from surfing lore.  For example, one frequent
suggestion {\bf Tim I need citations here} is that Brazilian surfers
such as Ferreira and Toledo have surfing skillsets that favour point
and beach break beaches {\bf Tim I need some help describing and
  defining these here}, those types of beach being readily accessible
to these surfers' native Brazil.  We now assess the null that
Brazilians perform equally well on all types of beaches.  To that end,
we create a reified surfer, ``Brazilian beachtype", whose strength
helps all Brazilian surfers when surfing on their (putative) preferred
beachtype, analogously to the ``home ground monster" who helps the
home team as per equation~\ref{home_ground_advantage}.  Here we modify
the likelihood function only when a single Brazilian is in the water,
to avoid possible interaction with the compatriot issue discussed in
section~\ref{compatriot}.  The null would be that this reified entity
has zero strength.  We fail to reject this null (differential support
= 0.035; $p=0.79$) and conclude that there is no evidence to suggest
that Brazilians' performce is better on their preferred beach types
than one would expect by chance.  We may go further using the method
of support: not only is the maximum likelihood estimate for Brazilian
beachtype very small (about 0.002), but in addition we may reject the
the proposal that the strength of this entity exceeds about 0.025
(Figure~\ref{brazilian_beachtype_support}).

@@

\section{Compatriot selection}

One measure of the rising popularity of surfing is its inclusion in
the 2020 Summer Olympics~\cite{tulloch2019}, and it is natural to ask
whether this would affect the competition in some way~\cite{ho2021}.
One suggestion would be that competitors would become more focussed on
beating compatriots than surfers from other countries, the thinking
being that admission to a country's Olympic squad would be of more
importance to a competitor than their result in wct.  One might expect
competitors to value their ranking when restricted to within their
country more than their ranking over all of wct.  If this is the case
then we would expect competitors to change their behaviour if they are
in the water with another surfer of the same nationality.
Specifically, if we have three surfers in the water of whom two are of
one nationality and the third of another, we would expect the
compatriots to adopt a style calculated to beat their countryman,
possibly at the expense of performance relative to the odd man out.

@@

Consider, as an example, the case of Smith, Buchan, Freestone.  Buchan
and Freestone are Australian, and Jordy Smith is South African.  It is
reasonable to ask whether Buchan and Freestone, who are in direct
competition with one another for a place in the Australian Olympic
surfing squad, would be more motivated to beating each other than to
come ahead of Smith.  This would be a rational response, but arguably
noncompetitive.  How would such a strategy manifest itself in
scorelines?

Reified Bradley-Terry technique furnishes an intuitive statistical
test for such a mechanism which controls for competitors' strengths.
Noting that the compatriots effectively place themselves at a {\em
  disadvantage} when compared with the odd man out, we introduce a
compatriot monster whose strength $p_C$ helps the odd man out.
Writing $p_1, p_2$ for the strengths of the compatriots and $q$ for
the odd man out, we construct likelihood functions for the six
possible orderings when two compatriots and an odd man out are in the
water at the same time in Table~\ref{likelihoodcompatriot} (if we have
either three different nationalities or three identical nationalities,
we revert to the standard Plackett-Luce likelihood).  Note that all
six likelihood functions in table~\ref{likelihoodcompatriot} are
informative about $p_1,p_2,q$ in addition to being informative about
the object of inference, $p_C$.

@@

We can now apply the Method of Support to the hypothesis that $p_C=0$,
that is, that this noncompetitive behaviour is absent.  We find that
the maximum likelihood estimate for $p_C$ is about 0.015, weaker than
the ``real'' surfers, but the support for a nonzero $p_C$ is about 1.8
[asymptotic $p=0.28$], short of the two units of support required for
rejection of our null.  We conclude that there is no strong evidence
to suggest that such noncompetitive behaviour occurred, at least not
from the order statistics we considered.

@@

{\bf\large Tim we need to discuss this}

>>>>

Figure~\ref{piechartmax} shows the maximum likelihood estimate for the
23 competitors' strengths, which appears to show a wide range from
Florence at about 17.5\% down to Coffin at about 1.5\%.  The {\tt
  hyper2} software~\cite{hankin2017} may be used to formally assess
one plausible null: that the competitors all have equal skill, that is
$H_0\colon p_\mathrm{Andino} = p_\mathrm{Bourez}=\ldots=
p_\mathrm{Wright}=\frac{1}{23}$, and any differences in placings are
due to random variation.  The Method of Support \cite{edwards1992}
rejects the null (in favour of the evaluate) at $p=4\times 10^{-5}$.
There is thus strong evidence to suggest that the competitors are
indeed of different skills and the result actually reflects genuine
differences in abilities, rather than the result of random chance.
Alternatively, we may view the dataset as being sufficiently
informative to reveal actual differences in surfing ability.  Observe
that the competitors' points (that is, the final column of
table~\ref{resultstable}) cannot be used for this type of
probabalistic analysis: the points awarded are arbitrary and changing
the points system would change the points scored but not affect the
competitors' behaviour.

@@

The statistical method described here can test many different
suggestions drawn from surfing lore.  For example, one frequent
suggestion~\cite{burgess2020,ho2021} is that Brazilian surfers such as
Ferreira and Toledo tend to have surfing skillsets that favour point
and beach break beaches {\bf Tim I need some help describing and
  defining these here}, those types of beach being readily accessible
to these surfers' native Brazil.  We now assess the null that
Brazilians perform equally well on all types of beaches.  To that end,
we create a reified surfer, ``Brazilian beachtype", whose strength
helps all Brazilian surfers when surfing on their (putative) preferred
beach type, an analogue of the ``home ground monster" who helps the
home team as per equation~\ref{home_ground_advantage}.  Here we modify
the likelihood function only when a single Brazilian is in the water,
to avoid possible interaction with the compatriot issue discussed in
section~\ref{compatriot}.  The null would be that this reified entity
has zero strength.  We fail to reject this null (differential support
= 0.035; $p=0.79$) and conclude that there is no evidence to suggest
that Brazilians' performce is better on their preferred beach types
than one would expect by chance.  We may go further using the method
of support: not only is the maximum likelihood estimate for Brazilian
beachtype very small (about 0.002), but in addition we may reject the
the proposal that the strength of this entity exceeds about 0.025
(Figure~\ref{brazilian_beachtype_support}).

@@

\section{Differential \label{compatriot} behaviour toward compatriotic competitors}

One measure of the rising popularity of surfing is its inclusion in
the 2020 Summer Olympics~\cite{tulloch2019}, and it is natural to ask
whether this would affect the competition in some way~\cite{ho2021}.
One suggestion would be that competitors become more focussed on
beating compatriots than surfers from other countries, the thinking
being that admission to a country's Olympic squad would be of more
importance to a competitor than their overall standing in the
tournament.  One might expect competitors to value their ranking when
restricted to within their country more than their international
ranking.  If this is the case then we would expect competitors to
change their behaviour if they are in the water with another surfer of
the same nationality.  Specifically, if we have three surfers in the
water of whom two are of one nationality and the third of another, we
would expect the compatriots to adopt a style calculated to beat their
countryman, possibly at the expense of performance relative to the odd
man out.

@@

Consider, as an example, the case of Smith, Buchan, Freestone who were
simultaneously in the water at Bells Beach, Australia.  Now, Buchan
and Freestone are Australian, and Jordy Smith is South African.  It is
reasonable to ask whether Buchan and Freestone, who are in direct
competition with one another for a place in the Australian Olympic
surfing squad, would be more motivated to beating each other than to
come ahead of Smith.  This would be a rational response, but arguably
noncompetitive.  How would such a strategy manifest itself in
scorelines?

Reified Bradley-Terry technique furnishes an intuitive statistical
test for mechanisms that control for competitors' strengths.  Noting
that the compatriots effectively place themselves at a {\em
  disadvantage} when compared with the odd man out, we introduce a
compatriot monster whose strength $p_C$ helps the odd man out.
Writing $p_1, p_2$ for the strengths of the compatriots and $q$ for
the odd man out, we construct likelihood functions for the six
possible orderings when two compatriots and an odd man out are in the
water at the same time in Table~\ref{likelihoodcompatriot} (if we have
either three different nationalities or three identical nationalities,
we revert to the standard Plackett-Luce likelihood).  Note that all
six likelihood functions in table~\ref{likelihoodcompatriot} are
informative about $p_1,p_2,q$ in addition to being informative about
the object of inference, $p_C$.

@@

We can now apply the Method of Support to the hypothesis that $p_C=0$,
that is, that such noncompetitive behaviour is absent.  We find that
the maximum likelihood estimate for $p_C$ is about 0.015, weaker than
the ``real'' surfers, but the support for a nonzero $p_C$ is about 1.8
[asymptotic $p=0.28$], slightly short of the two units of support
required for rejection of our null.  We conclude that there is
evidence, but no strong evidence, to suggest that such noncompetitive
behaviour occurred---at least not from the order statistics we
considered.

@@

We analysed results from the 2018 World Surf League using reified
Bradley-Terry likelihoods to assess a number of hypotheses of interest
to the surfing community.  Reified Bradley-Terry allows the use of
formal statistical methods in ways that analysing points scored does
not.  For example, we can be confident that the results reflect
differences in skill (rather than random variation); there is no
statistical difference in strength between the top two officially
ranked competitors, Medina and Ferreira; but we may reject the
hypothesis that Medina and Wilson are equal in strength.

We discuss and assess the ``Brazilian storm'' phenomenon using reified
entities and show that there is no strong evidence to support the
supposed Brazilian preference for point and beach break beaches.  We
also find no evidence to suggest that surfing's being included in the
2020 Summer Olympics results in modified strategies to bolster
intranational ranking at the expense WCT performance.

Further work might include  {\bf\large Tim we should discuss}
